---
title: "Gov 50 Final Project"
author: "Jack Holland"
description: "Analyzing swing voting behavior in CT"
output:
  distill::distill_article:
    self_contained: false
---



Introduction
---
___

What marks the difference between a swing voter and a strictly partisan one? Are there certain demographic variables that put people at a higher propensity to change their vote choice? Through data visualization and linear regression analysis, I plan to find answers to this phenomenon, as well as model novel ways to demonstrate swing voter determinants through data.

The scope of this study will be focused on my hometown of Fairfield, Connecticut. My hometown and county have historically entertained high rates of ticket splitting and a strong affinity for D-R reversion down ballot. However, both of the like are quickly diversifying and the phenomenon increasingly varies by voting precinct, evidenced by a number of state house elections flipping toward Democrats in the 2020 elections.

My analysis will come from the 2016 and 2020 Presidential Elections. 2020 Census American Community Service estimates (ACS) will be used to overlay the demographic data under which voting returns are analyzed. Raw numerical voting swings will define the incidence 'swing voters', analyzed at the precinct level. My hometown has an interesting slew of precincts - a semi-rural one, a highly urbanized one, a college campus, etc. Underneath the hood of presidential swings, demographics tell a compelling story.

I hypothesize that the swing voters of today are highly suburban, majority white, and have high levels of college attainment. These types of voters famously flocked toward Democrats in 2020, but it's not entirely clear that they have pledged their support toward down ballot Democrats just yet.

Why is this data interesting to me? Across the country, the incidence of swing and split-ticket voters is often reported to be waning. The Cook Political Report, for example, entitled its 2020 Presidential post-op article, "Lots of Turnout, Little Ticket Splitting", framing it is a moral conundrum that voters were likely to support their party in a straight-ticket fashion, through successive elections. I'm seeking to clarify preconceptions about the kind of people who change their vote, and what it means for future campaigns in my hometown.

---

```{r, echo=FALSE}
library(tidyr)
library(gov50data)
library(dplyr)
library(ggplot2)
library(htmltools)
library(patchwork)
```
The Data
---
___

My elections data is primarily sourced form the Harvard Dataverse project, a database used across the social sciences and not limited to Harvard affiliates. It includes two csv files: the 2016 and 2020 presidential elections. The data sets are merged by precinct to give a full record of their voting history. The demographic data used in this study is sourced from the American Community Survey, a study conducted by US Census which took place in 2020, not long before the Presidential Election. The demographic components will serve as the independent variable for regression analysis.

Though the elections data were sourced similarly, the data presented in each of the csv files was not entirely comparable, and required a good deal of tidying and mutating to reach a state in which the returns could be joined.

Coding voters as 'partisans' versus 'swing voters' is a tedious endeavor, especially when comparing between more than two political contests. Precinct data can give us a view of a specific population, but making generalizations down to the individual voter is impossible, and not the point of the study. Instead, I am looking to observe voting determinants within the context of a town and its distinct communities.


```{r}
#Biden 2020 Numbers
precincts2020 <- read.csv("/Users/jackholland/Downloads/School/GOV50/Datasets/2020precinct/PRESIDENT_precinct_general.csv")

#Clinton 2016 Numbers
pres16 <- read.csv("/Users/jackholland/Downloads/School/GOV50/Datasets/2016-precinct-president.csv")
pres16ct <- pres16 |>
  filter(state == "Connecticut") |>
  select(county_name, county_fips, county_ansi, jurisdiction, precinct, candidate_normalized, writein, party, votes)
```

```{r}
#ACS estimates by precinct
acs2020 <- read.csv("/Users/jackholland/Downloads/ACS2020.csv") |>
  select(Label..Grouping., ZCTA5.06824..Total..Estimate, ZCTA5.06825..Total..Estimate, ZCTA5.06890..Total..Estimate) |>
  rename(
    Demographics = Label..Grouping.,
    "06824" = ZCTA5.06824..Total..Estimate,
    "06825" = ZCTA5.06825..Total..Estimate,
    "06890" = ZCTA5.06890..Total..Estimate
  ) |>
  slice(12:25, 37:53)
head(acs2020)
```

___

```{r}
#Tidying election datasets for plotting purposes
clinton_support <- pres16ct |>
  mutate(cand_edit = if_else(candidate_normalized %in% c("trump", "clinton"), candidate_normalized, "Third Party")) |>
  filter(jurisdiction == "Fairfield")

biden_retention <- precincts2020 |>
  filter(
    state_po == "CT",
    jurisdiction_name == "FAIRFIELD",
    !is.na(party_simplified)
  ) |>
  mutate(party_simplest = if_else(party_simplified %in% c("DEMOCRAT", "REPUBLICAN"), party_simplified, "OTHER"))
```


Firstly, a look at the town wide election results in 2016 and 2020:

```{r, echo = FALSE}
clinton_plot <- clinton_support |>
  ggplot(aes(x = cand_edit, y = votes, fill = cand_edit)) +
  geom_bar(stat = "identity") +
  labs(
    title = "2016 President (Fairfield, CT)",
    x = "Candidate",
    y = "Vote Total"
  ) +
  labs(
        fill = "Candidate"
  ) +
  scale_x_discrete(labels = c(clinton = "Clinton", "trump" = "Trump")) +
  scale_fill_manual(values = c("Third Party" = "darkgoldenrod", "clinton" = "steelblue1", "trump" = "indianred1"),
                    labels = c("trump" = "Trump", "clinton" = "Clinton"))


biden_plot <- biden_retention |>
  ggplot(aes(x = party_simplest, y = votes, fill = party_simplest)) +
  geom_bar(stat = "identity") +
  labs(
    title = "2020 President (Fairfield, CT)",
    x = "Candidate",
    y = "Vote Total",
    fill = "Candidate"
  ) +
  scale_x_discrete(labels = c("OTHER" = "Third Party", "DEMOCRAT" = "Biden", "REPUBLICAN" = "Trump")) +
  scale_fill_manual(values = c("OTHER" = "darkgoldenrod", "DEMOCRAT" = "steelblue1", "REPUBLICAN" = "indianred1"),
                    labels = c("OTHER" = "Third Party", "DEMOCRAT" = "Biden", "REPUBLICAN" = "Trump")) +
  facet_wrap(~ "Candidate", ncol = 2)

clinton_plot + biden_plot
```

```{r, echo= FALSE, results='hide'}
#Renaming data and column names for joining
clinton_support <- clinton_support |>
  mutate(precinct_amended = precinct)

biden_retention <- biden_retention |>
  mutate(precinct = case_when(
    precinct == "DISTRICT 1-34-DWIGHT SCHOOL" ~ "Dwight School",
    precinct == "DISTRICT 10-32-MILL HILL SCHOOL" ~ "Mill Hill School",
    precinct == "DISTRICT 2-34-ST PIUS SCHOOL" ~ "St Pius School",
    precinct == "DISTRICT 4-33-STRATFIELD SCHOOL" ~ "Stratfield School",
    precinct == "DISTRICT 5-33-FAIRFIELD WARDE HIGH SCHOOL" ~ "Fairfield Warde High School",
    precinct == "DISTRICT 6-33-MCKINLEY SCHOOL" ~ "Mckinley School",
    precinct == "DISTRICT 7-33-HOLLAND HILL SCHOOL" ~ "Holland Hill School",
    precinct == "DISTRICT 8-32-FAIRFIELD LUDLOWE HIGH SCHOOL" ~ "Fairfield Ludlowe High School",
    precinct == "DISTRICT 9-32-SHERMAN SCHOOL" ~ "Sherman School",
    TRUE ~ "Fairfield Woods Middle School"))
```


```{r, results='hide'}
#Now, pivoting the tables to compare swing data
clinton_returns <- clinton_support |>
  group_by(precinct, cand_edit) |>
  summarize(vote_totals = sum(votes)) |>
  pivot_wider(
    names_from = cand_edit,
    values_from = vote_totals
  )

biden_returns <- biden_retention |>
  group_by(precinct, candidate) |>
  summarize(vote_totals = sum(votes)) |>
  pivot_wider(
    names_from = candidate,
    values_from = vote_totals
  )
```


```{r}
#Merging the Datasets...
composite_returns <- clinton_returns |>
  left_join(biden_returns, by = "precinct") |>
  select(precinct, clinton, trump, 'DONALD J TRUMP', 'JOSEPH R BIDEN') |>
  rename(
    "Precinct" = precinct,
    "Clinton16" = clinton,
    "Trump16" = trump,
    "Trump20" = 'DONALD J TRUMP',
    "Biden20" = 'JOSEPH R BIDEN'
  ) |>
  mutate(
    turnout_16 = (Clinton16 + Trump16),
    turnout_20 = (Biden20 + Trump20),
    dmarg_16 = ((Clinton16-Trump16)/turnout_16),
    dmarg_20 = ((Biden20-Trump20)/turnout_20),
    comp_swing = (dmarg_20-dmarg_16)*100)
(knitr::kable(head(composite_returns)))
```

And finally, our dependent variable... presidential swings by precinct:

```{r}
composite_plot <- composite_returns |>
  filter(!is.na(comp_swing)) |>
  ggplot(
    aes(x=reorder(Precinct, comp_swing),
        y =comp_swing)
  ) +
  geom_bar(stat = 'identity', mapping = aes(fill = comp_swing)) +
  labs(
    title = "2016-2020 Presidential Swings",
    x = "Precinct (School District)",
    y = "Two Way Raw Swing (D-R), in percentage points",
    fill = "D Swing (%)")

print(composite_plot)

```

___

The Results
---

The hard part of comparing election returns with ACS data is that the Census Bureau offers comprehensive location filters, but very few precinct or electoral boundary options. Thus, we have to water down the precinct returns into the three ZIP codes they are nested inside. Such limits the available points of comparison between districts, but doesn't mess with the calculations, by virtue of how the districts are nested. Below, the two datasets are merged:

```{r}
composite_final <- composite_returns |>
  mutate(zip_code = case_when(
    Precinct == "Mill Hill School" ~ 06890,
    Precinct == "Stratfield School" ~ 06825,
    Precinct == "Fairfield Woods Middle School" ~ 06825,
    Precinct == "Fairfield Warde High School" ~ 06825,
    TRUE ~ 06824)) |>
  ungroup() |>
  select(comp_swing, zip_code) |>
  group_by(zip_code) |>
  summarise(mean_comp_swing = mean(comp_swing, na.rm = TRUE))

composite_final$zip_code <- as.character(composite_final$zip_code)
composite_final$zip_code <- sprintf("%05d", as.numeric(composite_final$zip_code))
head(composite_final)
```
    
```{r}
acs_crosstab <- acs2020 |>
  pivot_longer(
    cols = `06824`:`06890`,
    values_to = "totals",
    names_to = "zip_code"
  ) |>
  group_by(zip_code)
head(acs_crosstab)
acs_crosstab$zip_code <- as.character(acs_crosstab$zip_code)
```

Now, to merge the demographic and voting data sets:

```{r}
voting_demographics <-
  left_join(acs_crosstab, composite_final, by = "zip_code") |>
  ungroup() |>
  filter(!totals == "")
head(voting_demographics)
```
```{r}
str(composite_final)
regression <- lm(mean_comp_swing ~ Demographics, data = voting_demographics)

modelsummary::modelsummary(regression,
statistic = c("s.e. = {std.error}",
"p = {p.value}"),
gof_map = c("nobs", "r.squared", "adj.r.squared"))
```



Conclusion
---

Swing voting, on a minute level, is indeed spurred by a specific slew of demographic factors.


